Sequential rank agreement methods for comparison of lists

### 2015-08-05 latex/paper.Rnw
Fixing application section and correcting minor spelling mistakes

\begin{abstract}
  The comparison of alternative rankings of a set of items is a
  general and prominent task in applied statistics. Predictor
  variables are ranked according to magnitude of association with an
  outcome, prediction models rank subjects according to the
  personalized risk of an event, and genetic studies rank genes
  according to their difference in gene expression levels. This
  article constructs measures of the agreement of two or more ordered
  lists. We use the standard deviation of the ranks to define a
  measure of agreement that both provides an intuitive interpretation
  and can be applied to any number of lists even if some or all are
  incomplete or censored. The main idea is then a permutation based
  and graphically approach to assess the sequential changes of
  agreement as a function of the depth of the lists. This may help to
  indentify change-points in the agreement of the lists. The
  usefulness of these tools are illustrated using gene rankings, and
  using data from two Danish ovarian cancer studies on where we assess
  the within and between agreement of different statistical classification
  methods.
\end{abstract}

@@

We now describe what is exemplified in Panel (c) of Table
\ref{tab:example} and how it can be used to define \emph{sequential
  rank agreement}. For an integer $1\le d\le P$ we define the unique
set of items found by merging the first $d$ elements of each of the
$L$ lists, i.e., the set of items ranked less than or equal to $d$ in
any of the lists:
\begin{equation}
S_d = \{R_l^{-1}(r) ; r \leq d, l = 1, \ldots, L \}.
\end{equation}
The \emph{sequential rank agreement} is the pooled standard deviation
of the items found in the set $S_d$:
\begin{equation}
\textrm{SRA}(d)= \sqrt{\frac{\sum_{\{p \in
      S_d\}}(L-1)A(X_p)^2}{(L-1)|S_d|}}, \label{def:sra}
\end{equation}
where $|S_d|$ refers to the cardinaly of the set $S_d$. Values of
\textrm{SRA} close to zero suggests that the lists agree while larger
values suggests disagreement. If the ranked lists are identical then
the value of SRA will be zero for all depths $d$. The sequential rank
agreement can be interpreted as the average distance of the individual
rankings of the lists from the average ranking for each of the items
we have seen until depth $d$. Note that the terms that are part of
\eqref{def:sra} are not independent since each list contains the ranks
from 1 to $P$ exactly once. Hence we simply use the pooled standard
deviation as a measure of the average distance between rankings.

@@

With fully observed lists we can plot the sequential rank agreement
\eqref{def:sra} as a function of depth $d$. An example is seen in top
panels of Figure~\ref{fig:example1} where four analysis methods were
used to rank 3051 gene expression values measured on 38 tumor mRNA
samples in order to classify between acute lymphoblastic leukemia
(ALL) and acute myeloid leukemia (AML)
\citep{Golub1999}. Preprocessing of the gene expression data was done
as described in \citet{Dudoit2002} and the four different analysis
approaces were: marginal two-sample $t$ tests, marginal logistic
regression analyses, logistic regression eleastic net
\citep{friedman2010regularization}, and marginal maximum information
content correlations (MIC) \citep{Reshef2011}. For the first two
methods, the genes were ranked according to $p$ value, for logistic
regression the genes were ordered by size of the corresponding
coefficients (after standardization), and MIC was ordered by
correlation which resulted in the top rankings seen in Table~\ref{tab1}.

@@

The average overlap is widely used in comparisons of (two) ranked
lists \citep{Fagin2003,Webber2010} and the idea behind the average
overlap closely resembles the sequential rank agreement. In this
section we present an extention of the average overlap to more that
two lists and compare this to sequential rank agreement.

@@

The black lines in Figure~\ref{fig:case1} show the average overlap and
sequential rank agreement for these data. It is clear from both plots
that there is perfect agreement towards the top of the ranked lists:
the average overlap is 1 and the sequential rank agreement is zero.
However, if we remove those two items from the data (i.e.,
gene/predictor 2124 and 896) and redo the analyses then we get
substantially different curves for the average overlap but not for
sequential rank agreement (red lines in Figure~\ref{fig:case1}). For
the sequential rank agreement we get roughly the same estimate of
agreement from rank 3 as we did from the full dataset. This indicates
that sequential rank is more robust against small pertubations of the
data. For the average overlap the curves completely change when these
two items are removed and which would lead to a different conclusion
about the agreement among the items from rank 3 downwards.

@@

In both scenarios we use the MALOVA data to train the statistical models, and the two types of agreements are assessed with respect to perturbations of the training data in the following manner. We repeatedly draw a sub-sample (without replication) consisting of 90\% of the MALOVA observations and train the four models on the sub-sample. Each of these models produces a ranking of the 5000 predictors according to their importance with respect to discriminating between  the tumor types, and the agreement of these rankings across the different sub-samples are used to assess the stability of the importance of the predictors. For the Random Forest classifier the predictors are ranked according to the Gini index, while for the logistic Lasso and Ridge regression models we order the predictors by absolute magnitude of the estimated regression coefficients. For the PLS-DA model the importance of the predictors is based on a weighted sum of the absolute coefficients where the weights are proportional to the reduction in the sums of squares across the components. To assess the stability of the individual risk predictions we apply the predictors from the DACOVA data set to each of the models. These predicted probabilities are then ranked in decreasing order, and the agreement of the risk predictions are similarly obtained by the sequential rank agreement across the different sub-samples of the training data. We use 1000 iterations for the sub-sampling procedure.

All four models depend on tuning parameters. We estimate these by a 20-times repeated 5-fold internal cross-validation on the training data in each iteration. The implementation of Lasso and Ridge regression in the \texttt{glmnet} package offers three different cross-validated optimization criteria for the penalty parameter: total deviance, classification accuracy and Area Under ROC. We apply all three criteria to our data to investigate their effect on the agreement.

@@

Figure \ref{fig:app1} (left) shows the sequential rank agreement applied to the estimated risk probabilities of the $113$ patients in the DACOVA data set obtained from the $1000$ sub-samples of the MALOVA data, and Figure \ref{fig:app1} (right) shows the sequential rank agreement of the estimated importance of the 5000 predictors. For clarity we only show the agreement for the latter up to list depth 600 where all the curves have approximately become stable. We note here that the predictor importance estimated with the logistic Lasso regression model is capable of producing censored lists depending on the value of the penalty parameter. With the deviance criterion the median number of non-zero coefficients was 33 (range 18 to 52). In order to calculate the sequential rank agreement we therefore applied Algorithm \ref{sra-algorithm} with $B = 1000$ and $P=5000$. The effect of the censoring caused by the sparsity of the Lasso is clearly visible in the agreement curve for the importance of the predictors.

For most of the sequential rank agreement curves we see as expected that they start low followed by an increase until they approximately become constant.  This has the interpretation that the classification methods show higher agreement with respect to the highest ranked items (the most high-risk patients and the most important predictors respectively) then for the items deeper in the lists across the different sub-samples. The change-point where the curves become approximately constant is the list depth where the ranks of the remaining items become almost random. The only exceptions are the predictor importance agreements using the logistic Ridge regression models and the risk probability agreement using the logistic Lasso regression with the AUC criterion which show higher disagreement in the top of the lists. The reason for this in the Lasso case is that the AUC criterion tends to favor a very high value of the penalty parameter causing only a single predictor to be selected in all of the 1000 iterations. This results in a lack of generalizability to the DACOVA data  giving rise to the higher disagreement in the predicted risk probabilities. In the extreme case where the penalty becomes so high that none of the predictors are selected by the Lasso, the sequential rank agreement for the predicted probabilities becomes undefined since all the ranks will be ties.

From Figure \ref{fig:app1} it can also be seen that some of the methods appear to be better at agreeing with respect to the predicted probabilities than at agreeing with respect to the ranking of the importance of the predictors and vice versa. Ridge regression performs very well for the risk predictions but is outperformed by PLS-DA for the predictor importances. On the other hand, PLS-DA appears to be similarly worse for the risk predictions compared to Ridge regression. The Random Forest method seems to perform equally well in both scenarios. Lasso regression (except with the AUC criterion as discussed previously) performs very well for the high-risk patients but very bad for the importance of the predictors. This reason for this is the high auto-correlation between the intensities in the mass spectra which leads to collinearity in the regression model. It is well-known that variable selection with the Lasso penalty does not preform very well when the predictors are highly correlated. The collinearity does, however, not affect the risk predictions, since it is not so important what specific variable that get selected from a group of highly correlated predictors when the purpose is risk predictions.

Finally, it would appear that Ridge regression with the AUC criterion for the penalty parameter achieves the best performance with respect to the stability of ranking the individual predicted risk probabilities. It must, however, be stressed that sequential rank agreement is only concerned with whether similar orderings of the patients is obtained under perturbations of the training data. It is not concerned with whether such an ordering is the correct one. This is similarly to the usual distinction between accuracy and precision. Looking at the distributions of the AUCs across the $1000$ iterations for each of the models with respect to the known class labels of the DACOVA data we see that the PLS-DA model attained the best AUC with a median value of $0.71$ (IQR = 0.07) while the Ridge regression with AUC criterion has a median AUC of $0.49$ (IQR = 0.02). This means that while Ridge regression with the AUC criterion achieves the highest rank agreement it performs similar to a random coin toss with respect to classification. Both concerns are of course important in practice and for this data set it appears that PLS-DA is the best method with respect to both concerns.

@@

The sequential rank agreement is extremely versatile. We have shown
that it can be used not only to compare ranked lists of items produced
from different samples/populations but that it also can be used to
study the ranks obtained from different analysis methods on the same
data as well as to evalaute the stability of the ranks from a single
method by bootstrapping the data repeatedly and comparing the ranks
obtained from the bootstrapped data.

@@

While the sequential rank agreement is primarily an exploratory tool
we have suggested two null hypotheses that can be used to evaluate the
sequential rank agreement obtained. Note that none of the two null
hypotheses are concerned with the actual ``true ranking'' but are
purely concerned with consistency/stability of the rankings among the
lists. As such we cannot determine if the rankings are good but only
whether they agree. The sequential rank agreemenet curve can be
compared visually to the curves obtained under either of the null
distributions and simple point-wise $p$-values can be obtained for
each depth by counting the number of sequential rank agreements under
the null hypothesis that is less than or equal to the observed rank
agreement. Two simple extensions can be pursued to make more formal
uniform tests: one would be to use a Kolmogorov-Smirnov-like test and
use the largest difference between the variance-weighted sequential
rank agreement curve and the mean null rank agreement curve as a test
statistic. Alternatively, a change-point analysis could be made on the
sequential rank agreement in order to determine the depths at which
there are ``jumps'' in the rank agreement. These jumps would
correspond to depths for which the agreement among the lists was
substantially worse and could serve as indicators for when the lists
no longer agree sufficiently satisfactory.

@@

We note that the sequential rank agreement is still marred by problems
that generally apply to ranking of items and/or
indiviuals. Collinearity in particular can be a huge problem when
bootstrapping data or when comparing different analysis methods. For
example, marginal analyses where each item is analyzed seperately will
assign similar ranks to two highly correlated predictors while methods
that provide a sparse solution such as the lasso will just rank one of
the two predictors high while the other might have a very low
rank. When the rankings from these two analysis methods are compared
then the agreement will differ simply because of the way correlated
predictors are handled. This is not a shortcoming of the sequential
rank agreement but is a problem general to all ranked lists.

>>>>

\begin{abstract}
  The comparison of alternative rankings of a set of items is a
  general and prominent task in applied statistics. Predictor
  variables are ranked according to magnitude of association with an
  outcome, prediction models rank subjects according to the
  personalized risk of an event, and genetic studies rank genes
  according to their difference in gene expression levels. This
  article constructs measures of the agreement of two or more ordered
  lists. We use the standard deviation of the ranks to define a
  measure of agreement that both provides an intuitive interpretation
  and can be applied to any number of lists even if some or all are
  incomplete or censored. The main idea is then a permutation based
  and graphically approach to assess the sequential changes of
  agreement as a function of the depth of the lists. This may help to
  identify change-points in the agreement of the lists. The
  usefulness of these tools are illustrated using gene rankings, and
  using data from two Danish ovarian cancer studies on where we assess
  the within and between agreement of different statistical classification
  methods.
\end{abstract}

@@

We now describe what is exemplified in Panel (c) of Table
\ref{tab:example} and how it can be used to define \emph{sequential
  rank agreement}. For an integer $1\le d\le P$ we define the unique
set of items found by merging the first $d$ elements of each of the
$L$ lists, i.e., the set of items ranked less than or equal to $d$ in
any of the lists:
\begin{equation}
S_d = \{R_l^{-1}(r) ; r \leq d, l = 1, \ldots, L \}.
\end{equation}
The \emph{sequential rank agreement} is the pooled standard deviation
of the items found in the set $S_d$:
\begin{equation}
\textrm{SRA}(d)= \sqrt{\frac{\sum_{\{p \in
      S_d\}}(L-1)A(X_p)^2}{(L-1)|S_d|}}, \label{def:sra}
\end{equation}
where $|S_d|$ refers to the cardinality of the set $S_d$. Values of
\textrm{SRA} close to zero suggests that the lists agree while larger
values suggests disagreement. If the ranked lists are identical then
the value of SRA will be zero for all depths $d$. The sequential rank
agreement can be interpreted as the average distance of the individual
rankings of the lists from the average ranking for each of the items
we have seen until depth $d$. Note that the terms that are part of
\eqref{def:sra} are not independent since each list contains the ranks
from 1 to $P$ exactly once. Hence we simply use the pooled standard
deviation as a measure of the average distance between rankings.

@@

With fully observed lists we can plot the sequential rank agreement
\eqref{def:sra} as a function of depth $d$. An example is seen in top
panels of Figure~\ref{fig:example1} where four analysis methods were
used to rank 3051 gene expression values measured on 38 tumor mRNA
samples in order to classify between acute lymphoblastic leukemia
(ALL) and acute myeloid leukemia (AML)
\citep{Golub1999}. Preprocessing of the gene expression data was done
as described in \citet{Dudoit2002} and the four different analysis
approaches were: marginal two-sample $t$ tests, marginal logistic
regression analyses, elastic net logistic regression 
\citep{friedman2010regularization}, and marginal maximum information
content correlations (MIC) \citep{Reshef2011}. For the first two
methods, the genes were ranked according to $p$ value, for logistic
regression the genes were ordered by size of the corresponding
coefficients (after standardization), and MIC was ordered by
correlation which resulted in the top rankings seen in Table~\ref{tab1}.

@@

The average overlap is widely used in comparisons of (two) ranked
lists \citep{Fagin2003,Webber2010} and the idea behind the average
overlap closely resembles the sequential rank agreement. In this
section we present an extension of the average overlap to more that
two lists and compare this to sequential rank agreement.

@@

The black lines in Figure~\ref{fig:case1} show the average overlap and
sequential rank agreement for these data. It is clear from both plots
that there is perfect agreement towards the top of the ranked lists:
the average overlap is 1 and the sequential rank agreement is zero.
However, if we remove those two items from the data (i.e.,
gene/predictor 2124 and 896) and redo the analyses then we get
substantially different curves for the average overlap but not for
sequential rank agreement (red lines in Figure~\ref{fig:case1}). For
the sequential rank agreement we get roughly the same estimate of
agreement from rank 3 as we did from the full dataset. This indicates
that sequential rank is more robust against small perturbations of the
data. For the average overlap the curves completely change when these
two items are removed and which would lead to a different conclusion
about the agreement among the items from rank 3 downwards.

@@

In both scenarios we use the MALOVA data to train the statistical models, and the two types of agreements are assessed with respect to perturbations of the training data in the following manner. We repeatedly draw a sub-sample (without replication) consisting of 90\% of the MALOVA observations and train the four models on the sub-sample. Each of the models produces a ranking of the 5000 predictors according to their importance with respect to discriminating between  the tumor types, and the agreement of these rankings across the different sub-samples are used to assess the stability of ranking the importance of the predictors. For the Random Forest classifier the predictors are ranked according to the Gini index, while for the logistic Lasso and Ridge regression models we order the predictors by absolute magnitude of the estimated regression coefficients. For the PLS-DA model the importance of the predictors is based on a weighted sum of the absolute coefficients where the weights are proportional to the reduction in the sums of squares across the components. To assess the stability of the individual risk predictions we apply the predictors from the DACOVA data set to each of the models. These predicted probabilities are then ranked in decreasing order, and the agreement of the risk predictions are similarly obtained by the sequential rank agreement applied across the different sub-samples of the training data. We use 1000 iterations for the sub-sampling procedure.

All four models depend on tuning parameters. These are estimated by 20 times repeated 5-fold cross-validation on the training data in each iteration. The implementation of Lasso and Ridge regression in the \texttt{glmnet} package offers three different cross-validated optimization criteria for the penalty parameter: total deviance, classification accuracy and Area Under ROC. We apply all three criteria to our data to investigate their effect on the agreement.

@@

Figure \ref{fig:app1} (left) shows the sequential rank agreement applied to the estimated risk probabilities of the $113$ patients in the DACOVA data set obtained from the $1000$ sub-samples of the MALOVA data, and Figure \ref{fig:app1} (right) shows the sequential rank agreement of the estimated importance of the 5000 predictors. For clarity we only show the agreement for the latter up to list depth 600 where all the curves have approximately become stable. We note here that the predictor importance estimated with the Lasso model is capable of producing censored lists depending on the value of the penalty parameter. With the deviance criterion the median number of non-zero coefficients was 33 (range 18 to 52). In order to calculate the sequential rank agreement we therefore applied Algorithm \ref{sra-algorithm} with $B = 1000$ and $P=5000$. The effect of the censoring caused by the sparsity of the Lasso is clearly visible in the agreement curves.

For most of the sequential rank agreement curves we see as expected that they start low followed by an increase until they approximately become constant.  This has the interpretation that the classification methods show higher agreement with respect to the highest ranked items (the most high-risk patients and the most important predictors respectively) then for the items deeper in the lists across the different sub-samples. The change-point where the curves become approximately constant is the list depth where the ranks of the remaining items become almost random. The only exceptions are the predictor importance agreements using the Ridge models and the risk probability agreement using the Lasso model with the AUC criterion which show higher disagreement in the top of the lists. The reason for this in the Lasso case is that optimizing with respect to the AUC criterion tends to favor a very high value of the penalty parameter causing only a single predictor to be selected in each of the 1000 iterations. This results in a lack of generalizability to the DACOVA data which gives rise to the higher disagreement in the predicted risk probabilities. In the extreme case where the penalty becomes so high that none of the predictors are selected by the Lasso, the sequential rank agreement for the predicted probabilities becomes undefined since all the ranks will be ties.

From Figure \ref{fig:app1} it can also be seen that some of the methods show better agreement with respect to the predicted probabilities than for ranking  the importance of the predictors and vice versa. Ridge regression performs very well for the risk predictions but is outperformed by PLS-DA for the predictor importances. On the other hand, PLS-DA appears to be similarly worse for the risk predictions compared to Ridge regression. The Random Forest method seems to perform equally well in both scenarios. Lasso regression (except with the AUC criterion as discussed previously) performs very well at ranking the high-risk patients but very poorly for the importance of the predictors. This reason for this is the high auto-correlation between the intensities in the mass spectra which leads to collinearity issues in the regression models. It is well-known that variable selection with the Lasso penalty does not preform very well when the predictors are highly correlated. This collinearity does, however, not affect the risk predictions, as seen on the left part of the figure, since it is not so important which specific variable that get selected from a group of highly correlated predictors when the purpose is risk predictions.

Finally, it would appear that Ridge regression with the AUC criterion for the penalty parameter achieves the best performance with respect to the stability of ranking the individual predicted risk probabilities. It must, however, be stressed that the sequential rank agreement is only concerned with whether similar orderings of the patients is obtained under perturbations of the training data. It is not concerned with whether such an ordering is the correct one. This is similar to the usual distinction between accuracy and precision. Looking at the distributions of the AUCs across the $1000$ iterations for each of the models with respect to the known class labels of the DACOVA data we see that the PLS-DA model attained the best AUC with a median value of $0.71$ (IQR = 0.07) while the Ridge regression with AUC criterion has a median AUC of $0.49$ (IQR = 0.02). This means that while Ridge regression optimized with respect to the AUC criterion achieves the best rank agreement it performs similar to a random coin toss with respect to classification on the DACOVA data. Both concerns are of course important in practice and for this data set it appears that PLS-DA is the overall best method taking both agreement and classification into account.

@@

The sequential rank agreement is extremely versatile. We have shown
that it can be used not only to compare ranked lists of items produced
from different samples/populations but that it also can be used to
study the ranks obtained from different analysis methods on the same
data as well as to evaluate the stability of the ranks from a single
method by bootstrapping the data repeatedly and comparing the ranks
obtained from the bootstrapped data.

@@

While the sequential rank agreement is primarily an exploratory tool
we have suggested two null hypotheses that can be used to evaluate the
sequential rank agreement obtained. Note that none of the two null
hypotheses are concerned with the actual ``true ranking'' but are
purely concerned with consistency/stability of the rankings among the
lists. As such we cannot determine if the rankings are good but only
whether they agree. The sequential rank agreement curve can be
compared visually to the curves obtained under either of the null
distributions and simple point-wise $p$-values can be obtained for
each depth by counting the number of sequential rank agreements under
the null hypothesis that is less than or equal to the observed rank
agreement. Two simple extensions can be pursued to make more formal
uniform tests: one would be to use a Kolmogorov-Smirnov-like test and
use the largest difference between the variance-weighted sequential
rank agreement curve and the mean null rank agreement curve as a test
statistic. Alternatively, a change-point analysis could be made on the
sequential rank agreement in order to determine the depths at which
there are ``jumps'' in the rank agreement. These jumps would
correspond to depths for which the agreement among the lists was
substantially worse and could serve as indicators for when the lists
no longer agree sufficiently satisfactory.

@@

We note that the sequential rank agreement is still marred by problems
that generally apply to ranking of items and/or
individuals. Collinearity in particular can be a huge problem when
bootstrapping data or when comparing different analysis methods. For
example, marginal analyses where each item is analyzed separately will
assign similar ranks to two highly correlated predictors while methods
that provide a sparse solution such as the lasso will just rank one of
the two predictors high while the other might have a very low
rank. When the rankings from these two analysis methods are compared
then the agreement will differ simply because of the way correlated
predictors are handled. This is not a shortcoming of the sequential
rank agreement but is a problem general to all ranked lists.
